The Evolution of Phenotypically Invariant Genetic Networks

### 2021-08-11 kalman_walk_paper.tex
minor edits

Although the other submatrices do not affect the phenotype
through the impulse response function $h(t)$,
they may affect individual fitness in other ways. \revpoint{2}{4}
For instance,
the ``unreachable'' subspaces cannot be affected by any of the possible variations
in input produced by varying $u(t)$.
However, if the nature of the input changed
(for instance, by perturbing a previously constant environmental variable),
differences in the unreachable components might lead to differences in phenotypes
and thus represent cryptic genetic variation.
Similarly, ``unobservable'' subspaces might represent aspects of phenotype
that are not constrained by selection,
but might become important if conditions change.
Additionally, all subspaces may affect how the system deals with noise.

@@

\paragraph{Haldane's rule}\hypertarget{rev1:5}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
\revpoint{1}{4}

Distribution of system coefficients across the autosomes and sex chromosomes does not alone guarantee Haldane's rule.
Under this model, Haldane's rule will only occur when there is sufficient genetic distance seperating parentals
and when the geometry of the optimal set produces $F_2$ hybrid breakdown. \revpoint{2}{1}

@@

\paragraph{The origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probably important
% but this one may be strong since happens in all networks at once
As classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms have been proposed
whereby Dobzhansky--Muller incompatibilities could originate
under neutral genetic drift \citep{lynch2000origin}
or stabilizing selection \citep{fierst}.
The same holds for the ``pathway model'' \citep{lindtke2015genetic},
which is closer to the situation here.
However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
% In light of the few known incompatibilities, 
This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
two processes that typically act much more rapidly than genetic drift.
Experimental approaches designed to study the effect of drift on speciation which induced population bottlenecks found no reproductive isolation in the majority tested pairs\citep{fry2009laboratory, white2020past}. These bottlenecks might not maintain enough genetic variation for systems to drift at an appreciable rate, however. \revpoint{1}{1}
%\jss{Did you also want to add something about bottlenecking large versus small popoulations?}
%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D. melanogaster} exhibit evidence of adaptative evolution \citep{tang2009evolution}.
%
% In contrast, the model we develop here suggests that even under strictly neutral conditions, 
% system drift can lead to speciation, at a rate fast enough 
% to play a substantial role in species formation across the tree of life. 
However, our calculations suggest that even under strictly neutral processes,
hybrid fitness breaks down as a function of genetic distance
rapidly enough to play a substantial role in species formation across the tree of life. 
This is consistent with broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation rates observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).
More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids, implicating the action of system drift.

Selection could dramatically accelerate speciation by essentially the same process --
for instance, \citet{mani1990mutational} showed that populations
developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
as different alleles fix in different populations. \linelabel{citemani}
Fluctuating selection \citep{bell2010fluctuating, barton2001role} is ongoing, so could further accelerate speciation.
Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments. This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.
In the case where fluctuations are synchronized, due to the size of the neutral system space, 
populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}

@@

\paragraph{Geometric models} \linelabel{quantnote}
Our results describe a continuum of possible systems,
described by their continuous coefficient matrices.
The arguments we make (e.g., Figure~\ref{fig:conceptual_fig}) are often geometric in nature,
and so it would be interesting to ask whether Fisher's geometric model
\citep[as in, e.g.,][]{martin2014fishers,fraisse2016genetics,simon2017coadapted}
might provide more quantitative, generalizable predictions.
It is also natural to ask about the distribution of effect sizes of individual mutations
in a population evolving under this model,
and what the distribituion of selection coefficients would be
(at equilibrium, perhaps).
These are questions that might be best answered by simulation,
and could provide much additional information -- for instance,
are alleles typically dominant in their effects on system coefficients or fitness?

>>>>

Although the other submatrices do not affect the phenotype
through the impulse response function $h(t)$,
they may affect individual fitness in other ways \revpoint{2}{4}
(which we don't consider here).
For instance,
the ``unreachable'' subspaces cannot be affected by any of the possible variations
in input produced by varying $u(t)$.
However, if the nature of the input changed
(for instance, by perturbing a previously constant environmental variable),
differences in the unreachable components might lead to differences in phenotypes
and thus represent cryptic genetic variation.
Similarly, ``unobservable'' subspaces might represent aspects of phenotype
that are not constrained by selection,
but might become important if conditions change.
Additionally, all subspaces may affect how the system deals with noise.

@@

\paragraph{Haldane's rule}\hypertarget{rev1:5}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, the contribution of sex chromosomes to the system coefficients of $F_1$ males
are effectively equivalent to those of purely autosomal $F_2$ hybrids,
which are less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
\revpoint{1}{4}
Note, however, that presence of genetic variation on sex chromosomes
does not alone guarantee Haldane's rule:
under this model, Haldane's rule will only occur when there is sufficient genetic distance separating parentals
and when the nature of genetic variation produces $F_2$ hybrid breakdown. \revpoint{2}{1}

@@

\paragraph{The origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probably important
% but this one may be strong since happens in all networks at once
As classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selected genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms have been proposed
whereby Dobzhansky--Muller incompatibilities could originate
under neutral genetic drift \citep{lynch2000origin}
or stabilizing selection \citep{fierst}.
The same holds for the ``pathway model'' \citep{lindtke2015genetic},
which is closer to the situation here.
However, previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
% Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
% In light of the few known incompatibilities, 
This has led some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence}
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual},
two processes that typically act much more rapidly than genetic drift.
Supporting this view, experimental approaches designed to study the effect of drift on speciation
through induced population bottlenecks
found no reproductive isolation in the majority of tested pairs \citep{fry2009laboratory, white2020past}. \revpoint{1}{1}
%and incompatibilities identified between \emph{Drosophila simulans} and \emph{D.~melanogaster} exhibit evidence of adaptive evolution \citep{tang2009evolution}.
However, our calculations suggest that even under strictly neutral processes,
hybrid fitness breaks down as a function of genetic distance
rapidly enough to play a substantial role in species formation across the tree of life. 
This is consistent with broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation rates observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).
More recently, \citet{sanchez2021widespread} observed widespread transcriptional misregulation in \emph{Drosophila} hybrids,
which might be due to the action of system drift.

Although the phrase ``system drift'' as we use it implies neutral changes,
it does not mean that these changes are induced purely by neutral genetic drift.
Selection could dramatically accelerate speciation by the same process --
i.e., \emph{genetic draft} \citep{gillespie2000genetic}.
For instance, \citet{mani1990mutational} showed that populations
developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
as different alleles fix in different populations. \linelabel{citemani}
Fluctuating selection has similarly been shown to accelerate speciation
in some models \citep{bell2010fluctuating,barton2001role}.
Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments.
This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.
In the case where fluctuations are synchronized, due to the size of the neutral system space, 
populations may arrive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}

@@

\paragraph{Geometric models} \linelabel{quantnote}
The model we work with describes a continuum of possible systems,
parameterized by continuous coefficient matrices.
The arguments we make (e.g., Figure~\ref{fig:conceptual_fig}) are often geometric in nature,
and so it would be interesting to ask whether Fisher's geometric model
might provide more quantitative, generalizable predictions,
as in for instance \citet{fraisse2016genetics,simon2017coadapted}.
\citet{martin2014fishers} argued that such an approximation would be often fruitful.
In the other direction,
it is also natural to ask about the distribution of effect sizes of segregating mutations
in a population evolving under this model with a realistic \emph{de novo} effect size distribution,
as well as what the distribution of fitness effects would be (both within and between populations).
These are questions that might be best answered by simulation,
and could provide much additional information -- for instance,
are alleles typically dominant in their effects on either system coefficients or fitness?
